---
title: "Data Analysis for 'BCMS Prefix Priming'"
date: today
format:
  typst:
    papersize: us-letter
    margin:
      x: 0.5in
      y: 0.5in
    fontsize: 10pt
    toc: false
    number-sections: false
execute:
  echo: false
  warning: false
  message: false
bibliography: references.bib
editor_options:
  chunk_output_type: console
---

```{r load-packages}
library(tidyverse)
library(janitor)
library(ggpubr)
library(tidyverse)
library(BayesFactor)
library(purrr)
library(brms)
library(ggdist)
library(knitr)
```


```{r utility-functions}
# User-defined function to read in PCIbex Farm results files
read.pcibex <- function(filepath, auto.colnames=TRUE, fun.col=function(col,cols){cols[cols==col]<-paste(col,"Ibex",sep=".");return(cols)}) {
  n.cols <- max(count.fields(filepath,sep=",",quote=NULL),na.rm=TRUE)
  if (auto.colnames){
    cols <- c()
    con <- file(filepath, "r")
    while ( TRUE ) {
      line <- readLines(con, n = 1, warn=FALSE)
      if ( length(line) == 0) {
        break
      }
      m <- regmatches(line,regexec("^# (\\d+)\\. (.+)\\.$",line))[[1]]
      if (length(m) == 3) {
        index <- as.numeric(m[2])
        value <- m[3]
        if (is.function(fun.col)){
         cols <- fun.col(value,cols)
        }
        cols[index] <- value
        if (index == n.cols){
          break
        }
      }
    }
    close(con)
    return(read.csv(filepath, comment.char="#", header=FALSE, col.names=cols))
  }
  else{
    return(read.csv(filepath, comment.char="#", header=FALSE, col.names=seq(1:n.cols)))
  }
}


subj_id <- function(df, cols = c("results_reception_time", "md5_hash_of_participant_s_ip_address"), label = "subj") {
  lookup <- df %>%
    distinct(across(all_of(cols))) %>%
    mutate(!!label := paste0("S[", row_number(), "]"))

  df %>% left_join(lookup, by = cols)
}

okabe_ito_palette <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442", 
                       "#0072B2", "#D55E00", "#CC79A7", "#000000")

```


```{r load-data}
results <- bind_rows(
  read.pcibex("bcms_priming_1.csv") %>% clean_names(),
  read.pcibex("bcms_priming_2.csv") %>% clean_names()
) %>% subj_id()

prime_duration_lookup <- results %>%
  filter(condition == "Exp",
         penn_element_type == "Audio",
         penn_element_name == "prime",
         parameter == "end") %>%
  transmute(
    subj,
    group,
    item_ident,
    order = order_number_of_item,
    prime_duration = as.double(value)
  ) %>%
  group_by(subj, group, item_ident, order) %>%
  slice_max(prime_duration, n = 1, with_ties = FALSE) %>%
  ungroup()

target_duration_lookup <- results %>%
  filter(condition == "Exp",
         penn_element_type == "Audio",
         penn_element_name == "target",
         parameter == "end") %>%
  transmute(
    subj,
    group,
    item_ident,
    order = order_number_of_item,
    target_duration = as.double(value)
  ) %>%
  group_by(subj, group, item_ident, order) %>%
  slice_max(target_duration, n = 1, with_ties = FALSE) %>%
  ungroup()

exp <- results %>% 
  filter(condition == "Exp") %>% 
  filter(penn_element_name == "answerPrime") %>% 
  select(subj, group, item_ident, prime, target, ptype, pword, key_prime, tword, key_target, target_expected_key, prime_rt, target_rt, iti, isi, sequence, trial_n, order=order_number_of_item) %>%
  left_join(prime_duration_lookup, by = c("subj", "group", "item_ident", "order")) %>%
  left_join(target_duration_lookup, by = c("subj", "group", "item_ident", "order")) %>%
  mutate(ptype = recode(ptype,
      "MS" = "Morpho-Semantic",
      "M"  = "Morpho-Opaque",
      "PM" = "Pseudo-Morphological",
      "P"  = "Phonological",
      "C"  = "Control"
    ))

practice <- results %>% 
  filter(label == "practice") %>% 
  filter(parameter  == "PressedKey") %>% 
  select(subj, prime=prime_soundfile, primekey=prime_rt, target=target_soundfile, targetkey=target_rt) 


filler <- results %>% 
  filter(condition == "Filler") %>% 
  filter(penn_element_name == "answerPrime") %>% 
  select(subj, group, item_ident, prime, target, ptype, pword, key_prime, tword, key_target, target_expected_key, prime_rt, target_rt, iti, isi, sequence, trial_n, order=order_number_of_item)

N_subj <- n_distinct(exp$subj)
```



```{r filler-accuracy}
# practice and filler accuracy and participant exclusion

filler_accuracy <- filler %>%
    mutate(p_acc = as.integer((pword == "word" & key_prime == "F") |
                            (pword == "nonword" & key_prime == "J")),
           t_acc = as.integer((tword == "word" & key_target == "F") |
                            (tword == "nonword" & key_target == "J"))) %>%
  group_by(subj) %>%
  summarize(
    p_acc_mean = mean(p_acc),
    t_acc_mean = mean(t_acc)
  ) 

bad_subjects <- filler_accuracy %>%
  filter(p_acc_mean < 0.75 | t_acc_mean < 0.75) %>% pull(subj)

n_bad_subjects <- length(bad_subjects)
```

## About this study

This report analyzes the status of Bosnian/Croatian/Montenegrin/Serbian (BCMS) prefixex in mental grammar using a priming experiment. The work is done by Maša Beslin, Predrag Kovačević, Stefan Ivanović and Utku Turk. The key question is whether lexical prefixes can be individuated as morphemes. We test this by examining whether **prime condition** affects target lexical-decision behavior, and whether that effect remains once we control for trial progression, prime duration, target duration, prime RT, and Inter-Stimulus Interval (ISI).


`r N_subj` students at the University of Novi Sad, Serbia completed an unmasked auditory continous lexical decision task. Sample size was determined via a priori power analysis to ensure 80% power following [@Vasishthetal2023]. `r n_bad_subjects` participant(s) were excluded due to filler accuracy falling below 75%.


Critical targets are high frequency simplex verbs in the masculine active participle form. Prime-target relations include: 

- morphologically and semantically related (predao [gave over] --> dao [gave]), 
- morphologically related but semantically opaque (prodao [sold] --> dao [gave]),
- pseudo-morphologically related (nadao [hoped] --> dao [gave]),
- phonologically related (vladao [ruled] --> dao [gave]),
- unrelated controls (preneo [carried over] --> dao [gave])

Participants are assigned to one of four experimental lists, each containing 30 critical item pairs: 5 per priming condition and 10 controls. Each target appears only once per list but is paired with different prime types across lists, enabling within-item comparisons while avoiding repetition effects. 

We include 200 filler pairs, which consist of 50% real words and 50% non-words (by making single-segment modifications to existing verbs while respecting BCMS phonotactics). Due to language-specific constraints, the experiment employs a balanced incomplete block design, which accommodates the fact that not all targets can be paired across all conditions, while each condition is equally represented and can be compared to controls [@BechhoferTamhane1981].

The descriptive plots below first show accuracy, then prime RTs, and then target RTs. The Bayesian model later in the report is fit only to trials with a **correct response to the target**, after the RT trimming steps described below. We incorporate random effects for participants and items to account for within-subject and within-item variability [@Baayen2008]. Our predictors include prime condition, prime duration, target duration, prime Response Time (RT), trial number, and Inter-Stimulus Interval (ISI). Since we are mainly interested in by-condition differences, we treat our numeral predictors as log-transformed and centered (z-scored). Our baselines then are the 'average' prime duration, target duration, prime RT, trial number, and ISI. 

We employed weakly informative priors as follows. The intercept prior is centered on 6.8 (approximately $e^{6.8} \approx 900$ ms), reflecting a plausible grand-mean log RT. All fixed-effect slopes receive a $\text{Normal}(0, 0.5)$ prior, which is skeptical of large effects but permits the data to overwhelm the prior. Random-effect standard deviations and the residual SD use an $\text{Exponential}(1)$ prior, which gently regularizes toward smaller variance components. Correlation matrices among random effects are assigned an $\text{LKJ}(2)$ prior, which mildly favors lower correlations and helps regularize the maximal random-effect structure.

| Parameter | Prior | Rationale |
|:----------|:------|:----------|
| Intercept | $\text{Normal}(6.8,\; 0.5)$ | Centers on ~900 ms mean RT in log space |
| Fixed-effect slopes ($\beta$) | $\text{Normal}(0,\; 0.5)$ | Weakly informative; skeptical of large effects |
| Random-effect SDs | $\text{Exponential}(1)$ | Regularizes toward smaller variance components |
| Residual SD ($\sigma$) | $\text{Exponential}(1)$ | Same regularization for observation-level noise |
| Random-effect correlations | $\text{LKJ}(2)$ | Mildly favors lower correlations to stabilize estimation |

: Prior specification for the Bayesian mixed-effects model. {#tbl-priors}

### Model specification

The response variable is **log-transformed target RT**. Prime condition enters the model as a five-level treatment-coded factor with **Control** as the reference level, yielding four contrasts that each compare one priming condition against Control. The remaining fixed effects are continuous nuisance covariates, all z-scored.

| Component | Specification |
|:----------|:--------------|
| Response | $\log(\text{target RT})$ |
| Fixed effects | Prime condition (treatment-coded, ref = Control), z log prime duration, z log target duration, z log prime RT, z trial number, z ISI |
| Contrasts | Morpho-Semantic vs Control, Morpho-Opaque vs Control, Pseudo-Morphological vs Control, Phonological vs Control |
| By-subject random effects | Intercept + all fixed-effect slopes |
| By-item random effects | Intercept + all fixed-effect slopes |
| Family | Gaussian |
| Chains / Iterations | 4 chains × 6 000 iterations (2 000 warmup) |
| Sampler settings | `adapt_delta` = 0.99, `max_treedepth` = 15 |
| Backend | CmdStanR |

: Model specification summary. The maximal random-effect structure follows @Baayen2008. {#tbl-model-spec}

## Results

### Accuracy

Accuracy is shown first because it gives the clearest screening view of whether any prime condition is associated with a speed-accuracy tradeoff. The y-axis is expressed as a percentage rather than a proportion.

```{r accuracy-plot, fig.width=8.5, fig.height=3, fig.cap="Prime and target accuracy by prime condition. Error bars show 95% confidence intervals. The dashed line marks the Control-condition mean."}
# prime and target accuracy

exp_accuracy <- exp %>%
    mutate(p_acc = as.integer((pword == "word" & key_prime == "F") |
                            (pword == "nonword" & key_prime == "J")),
           t_acc = as.integer((tword == "word" & key_target == "F") |
                            (tword == "nonword" & key_target == "J"))) %>%
  group_by(ptype) %>%
  summarize(
    p_acc_mean = mean(p_acc),
    p_acc_se = sd(p_acc)/sqrt(n()), 
    t_acc_mean = mean(t_acc),
    t_acc_se = sd(t_acc)/sqrt(n())
  ) 


exp_accuracy_long <- exp_accuracy %>%
  pivot_longer(
    cols = c(p_acc_mean, t_acc_mean),
    names_to = "measure",
    values_to = "mean"
  ) %>%
  mutate(
    se = ifelse(measure == "p_acc_mean", p_acc_se, t_acc_se),
    measure = ifelse(measure == "p_acc_mean", "Prime", "Target"),
    mean_pct = mean * 100,
    se_pct = se * 100
  ) %>%
  select(ptype, measure, mean_pct, se_pct)

accuracy_control_lines <- exp_accuracy_long %>%
  filter(ptype == "Control") %>%
  transmute(measure, control_mean_pct = mean_pct)

ggplot(exp_accuracy_long, aes(x = ptype, y = mean_pct)) +
  geom_hline(
    data = accuracy_control_lines,
    aes(yintercept = control_mean_pct),
    linewidth = 0.4,
    linetype = "dashed",
    color = "grey55"
  ) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = mean_pct - 1.96 * se_pct, ymax = mean_pct + 1.96 * se_pct),
                width = 0.2) +
  facet_wrap(~measure) +
  labs(x = "Prime Type", y = "Accuracy (%)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

```{r rt-exclusion}
# Trials with incorrect responses to targets were discarded
d1 <- exp %>%
    mutate(p_acc = as.integer((pword == "word" & key_prime == "F") |
                            (pword == "nonword" & key_prime == "J")),
           t_acc = as.integer((tword == "word" & key_target == "F") |
                            (tword == "nonword" & key_target == "J"))) %>%
  filter(t_acc == 1) 

pct_excluded_accuracy <- round((nrow(exp) - nrow(d1)) / nrow(exp) * 100, 2)

# All targets with outlier RTs
# (<250ms and >2000ms) were excluded, as well as the targets for which the prime had an outlier RT
# (<250ms and >2500ms).
d1 <- d1 %>% mutate(prime_rt = as.double(prime_rt), target_rt = as.double(target_rt))
d2 <- d1 %>% filter(prime_rt > 250 & prime_rt < 2000 & target_rt > 250 & target_rt < 2500)

pct_excluded_rt <- round((nrow(d1) - nrow(d2)) / nrow(d1) * 100, 2)

exp <- d2
```

Trials with incorrect responses to the target were discarded (`r pct_excluded_accuracy`% of the data). Subsequently, trials with outlier RTs (target RT < 250 ms or > 2500 ms; prime RT < 250 ms or > 2000 ms) were removed, excluding an additional `r pct_excluded_rt`% of the remaining data.


### Prime RTs

The next figure summarizes prime RTs after restricting the timing data to target-correct trials and trimming outliers. The dashed horizontal line shows the mean RT in the Control condition, so each condition can be read relative to that baseline.

```{r prime-rt-summary}
prime_rt_summary <- exp %>%
  group_by(ptype) %>%
  summarise(
    n = n(),
    mean_prime_rt = mean(prime_rt),
    se_prime_rt = sd(prime_rt) / sqrt(n()),
    .groups = "drop"
  )

kable(
  prime_rt_summary,
  digits = 2,
  col.names = c("Prime condition", "N", "Mean prime RT (ms)", "SE"),
  caption = "Prime RTs for target-correct trials after RT trimming."
)
```

```{r prime-rt-plot, fig.width=8.5, fig.height=3.6, fig.cap="Prime RTs by prime condition, using only target-correct trials after RT trimming. The dashed line marks the Control-condition mean."}
prime_rt_control_mean <- exp %>%
  filter(ptype == "Control") %>%
  summarise(control_mean = mean(prime_rt)) %>%
  pull(control_mean)

exp %>%
  ggplot(aes(x = ptype, y = prime_rt, color = ptype)) +
  geom_hline(yintercept = prime_rt_control_mean, linewidth = 0.4, linetype = "dashed", color = "grey55") +
  geom_boxplot(aes(fill = ptype), width = 0.56, alpha = 0.18, outlier.shape = NA, show.legend = FALSE) +
  geom_point(position = position_jitter(width = 0.08, height = 0), alpha = 0.12, size = 1.1, show.legend = FALSE) +
  scale_color_manual(values = okabe_ito_palette) +
  scale_fill_manual(values = okabe_ito_palette) +
  labs(x = "Prime condition", y = "Prime RT (ms)") +
  theme_classic(base_size = 12) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

```{r prime-rt-check}
# Simple model: do prime RTs differ by condition?
prime_rt_model <- lmerTest::lmer(
  log(prime_rt) ~ ptype + (1 | subj) + (1 | item_ident),
  data = exp
)

prime_rt_anova <- anova(prime_rt_model)
prime_rt_F <- round(prime_rt_anova$`F value`[1], 2)
prime_rt_df1 <- prime_rt_anova$NumDF[1]
prime_rt_df2 <- round(prime_rt_anova$DenDF[1], 1)
prime_rt_p <- prime_rt_anova$`Pr(>F)`[1]
prime_rt_p_fmt <- ifelse(prime_rt_p < 0.001, "< .001", sprintf("= %.3f", prime_rt_p))

# Per-condition summaries for inline use
prime_rt_by_cond <- exp %>%
  group_by(ptype) %>%
  summarise(
    m = round(mean(prime_rt)),
    se = round(sd(prime_rt) / sqrt(n())),
    .groups = "drop"
  )
prt <- function(cond) {
  row <- prime_rt_by_cond %>% filter(ptype == cond)
  paste0(row$m, " ms (SE = ", row$se, ")")
}
```

Descriptively, mean prime RTs were similar across conditions: Control `r prt("Control")`, Morpho-Semantic `r prt("Morpho-Semantic")`, Morpho-Opaque `r prt("Morpho-Opaque")`, Pseudo-Morphological `r prt("Pseudo-Morphological")`, and Phonological `r prt("Phonological")`. A simple linear mixed-effects model predicting log prime RT from prime condition with random intercepts for subjects and items showed `r ifelse(prime_rt_p < 0.05, "a significant", "no significant")` main effect of condition, *F*(`r prime_rt_df1`, `r prime_rt_df2`) = `r prime_rt_F`, *p* `r prime_rt_p_fmt`. `r ifelse(prime_rt_p < 0.05, "This suggests that prime processing difficulty varied across conditions, which motivates the inclusion of prime RT as a covariate in the target RT model.", "This indicates that participants did not find any prime condition systematically harder to process, supporting the interpretation that target RT differences reflect priming rather than prime difficulty.")` Regardless, we include log prime RT as a z-scored covariate in our main target RT model to account for by-trial variation in processing speed, ensuring that any condition effects on target RTs are not confounded with trial-level speed differences.


### Target RTs

Target RTs are plotted on the log scale because the model is fit to log-transformed target RTs. The dashed horizontal line again marks the Control-condition mean, so the figure lines up with the model coefficients that compare each condition to Control.

```{r target-rt-data}
ptype_levels <- c(
  "Control",
  "Morpho-Semantic",
  "Morpho-Opaque",
  "Pseudo-Morphological",
  "Phonological"
)

ptype_simple_levels <- c(
  "Control",
  "MorphoSemantic",
  "MorphoOpaque",
  "PseudoMorphological",
  "Phonological"
)

ptype_labels <- c(
  Control = "Control",
  MorphoSemantic = "Morpho-Semantic",
  MorphoOpaque = "Morpho-Opaque",
  PseudoMorphological = "Pseudo-Morphological",
  Phonological = "Phonological"
)

exp_model <- exp %>%
  mutate(
    subj = factor(subj),
    item_ident = factor(item_ident),
    trial_n = as.double(trial_n),
    isi = as.double(isi),
    prime_duration = as.double(prime_duration),
    target_duration = as.double(target_duration),
    ptype = factor(ptype, levels = ptype_levels),
    ptype_simple = factor(ptype_simple_levels[match(ptype, ptype_levels)], levels = ptype_simple_levels),
    ptype_label = factor(ptype_labels[as.character(ptype_simple)], levels = unname(ptype_labels)),
    log_target_rt = log(as.double(target_rt)),
    log_prime_rt = log(as.double(prime_rt)),
    log_prime_duration = log(prime_duration),
    log_target_duration = log(target_duration),
    z_trial_n = as.numeric(scale(trial_n)),
    z_log_prime_rt = as.numeric(scale(log_prime_rt)),
    z_log_prime_duration = as.numeric(scale(log_prime_duration)),
    z_log_target_duration = as.numeric(scale(log_target_duration)),
    z_isi = as.numeric(scale(isi))
  ) %>%
  select(
    subj,
    item_ident,
    ptype_simple,
    ptype_label,
    log_target_rt,
    z_trial_n,
    z_log_prime_duration,
    z_log_target_duration,
    z_log_prime_rt,
    z_isi
  ) %>%
  drop_na()

target_rt_summary <- exp_model %>%
  group_by(ptype_label) %>%
  summarise(
    n = n(),
    mean_log_rt = mean(log_target_rt),
    se_log_rt = sd(log_target_rt) / sqrt(n()),
    .groups = "drop"
  )

kable(
  target_rt_summary,
  digits = 2,
  col.names = c("Prime condition", "N", "Mean log RT", "SE"),
  caption = "Log-transformed target RTs by prime condition."
)
```
```{r target-rt-descriptives}
# Per-condition summaries on the raw ms scale for prose
target_rt_by_cond <- exp %>%
  mutate(ptype = factor(ptype, levels = c("Control", "Morpho-Semantic", "Morpho-Opaque", "Pseudo-Morphological", "Phonological"))) %>%
  group_by(ptype) %>%
  summarise(
    m = round(mean(target_rt)),
    se = round(sd(target_rt) / sqrt(n())),
    m_log = round(mean(log(target_rt)), 3),
    se_log = round(sd(log(target_rt)) / sqrt(n()), 3),
    .groups = "drop"
  )

trt <- function(cond) {
  row <- target_rt_by_cond %>% filter(ptype == cond)
  paste0(row$m, " ms (SE = ", row$se, "; log: ", row$m_log, ")")
}

# Fastest and slowest conditions
fastest_cond <- target_rt_by_cond %>% slice_min(m, n = 1) %>% pull(ptype) %>% as.character()
slowest_cond <- target_rt_by_cond %>% slice_max(m, n = 1) %>% pull(ptype) %>% as.character()
rt_range <- max(target_rt_by_cond$m) - min(target_rt_by_cond$m)

n_model_obs <- nrow(exp_model)
n_model_subj <- n_distinct(exp_model$subj)
n_model_items <- n_distinct(exp_model$item_ident)

# Differences from Control
control_m <- target_rt_by_cond %>% filter(ptype == "Control") %>% pull(m)
control_m_log <- target_rt_by_cond %>% filter(ptype == "Control") %>% pull(m_log)

diff_from_ctrl <- target_rt_by_cond %>%
  filter(ptype != "Control") %>%
  mutate(
    diff_ms = control_m - m,
    diff_log = round(control_m_log - m_log, 3)
  )

dfc <- function(cond) {
  row <- diff_from_ctrl %>% filter(ptype == cond)
  dir <- ifelse(row$diff_ms > 0, "faster", "slower")
  paste0(abs(row$diff_ms), " ms ", dir, " (Δ log = ", sprintf("%.3f", row$diff_log), ")")
}

# Conditions faster than control
faster_conds <- diff_from_ctrl %>% filter(diff_ms > 0) %>% pull(ptype) %>% as.character()
faster_conds_str <- paste(faster_conds, collapse = ", ")
```

Mean target RTs by condition were: Control `r trt("Control")`, Morpho-Semantic `r trt("Morpho-Semantic")`, Morpho-Opaque `r trt("Morpho-Opaque")`, Pseudo-Morphological `r trt("Pseudo-Morphological")`, and Phonological `r trt("Phonological")`. Numerically, `r fastest_cond` elicited the fastest responses and `r slowest_cond` the slowest, with a range of `r rt_range` ms across conditions.

Compared to the Control condition, all morphologically related conditions showed descriptively faster target responses: Morpho-Semantic was `r dfc("Morpho-Semantic")`, Morpho-Opaque was `r dfc("Morpho-Opaque")`, and Pseudo-Morphological was `r dfc("Pseudo-Morphological")`. The Phonological condition was `r dfc("Phonological")`. These descriptive differences are consistent with a facilitation effect for morphological priming, though formal inference requires the model reported below. After excluding trials with missing duration values, the modeling dataset contains `r n_model_obs` observations from `r n_model_subj` participants across `r n_model_items` items.


```{r target-rt-boxplot, fig.width=8.5, fig.height=3.6, fig.cap="Log-transformed target RTs by prime condition. Boxes show the condition distribution; dots show individual observations with low alpha. The dashed line marks the Control-condition mean."}
target_rt_control_mean <- exp_model %>%
  filter(ptype_label == "Control") %>%
  summarise(control_mean = mean(log_target_rt)) %>%
  pull(control_mean)

exp_model %>%
  ggplot(aes(x = ptype_label, y = log_target_rt, color = ptype_label)) +
  geom_hline(yintercept = target_rt_control_mean, linewidth = 0.4, linetype = "dashed", color = "grey55") +
  geom_boxplot(aes(fill = ptype_label), width = 0.56, alpha = 0.18, outlier.shape = NA, show.legend = FALSE) +
  geom_point(position = position_jitter(width = 0.08, height = 0), alpha = 0.12, size = 1.1, show.legend = FALSE) +
  scale_color_manual(values = okabe_ito_palette) +
  scale_fill_manual(values = okabe_ito_palette) +
  labs(x = "Prime condition", y = "Log target RT") +
  theme_classic(base_size = 12) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

## Statistical analysis

The Bayesian model below uses only the cleaned timing dataset: trials with a correct response to the target, plus RT trimming for both prime and target. The outcome is **log target RT**, and the predictors are prime condition, z-scored trial number, z-scored log prime duration, z-scored target duration, z-scored log prime RT, and z-scored ISI.

```{r target-rt-model, include=FALSE}
model_cores <- 4

target_rt_priors <- c(
  prior(normal(0, 0.5), class = "b"),
  prior(normal(6.8, 0.5), class = "Intercept"),
  prior(exponential(1), class = "sd"),
  prior(exponential(1), class = "sigma"),
  prior(lkj(2), class = "cor")
)

target_rt_brm <- brm(
  formula = bf(
    log_target_rt ~
      ptype_simple +
      z_trial_n +
      z_log_prime_duration +
      z_log_target_duration +
      z_log_prime_rt +
      z_isi +
      (1 + ptype_simple + z_trial_n + z_log_prime_duration + z_log_target_duration + z_log_prime_rt + z_isi | subj) +
      (1 + ptype_simple + z_trial_n + z_log_prime_duration + z_log_target_duration + z_log_prime_rt + z_isi | item_ident)
  ),
  data = exp_model,
  family = gaussian(),
  prior = target_rt_priors,
  chains = 4,
  cores = model_cores,
  iter = 6000,
  warmup = 2000,
  seed = 20260413,
  backend = "cmdstanr",
  control = list(adapt_delta = 0.99, max_treedepth = 15),
  refresh = 0,
  file = "target_rt_brm_maximal",
  file_refit = "on_change"
)

target_draws <- posterior::as_draws_df(target_rt_brm) %>%
  as_tibble()

pretty_target_term <- function(term) {
  case_match(
    term,
    "ptype_simpleMorphoSemantic" ~ "Morpho-Semantic vs Control",
    "ptype_simpleMorphoOpaque" ~ "Morpho-Opaque vs Control",
    "ptype_simplePseudoMorphological" ~ "Pseudo-Morphological vs Control",
    "ptype_simplePhonological" ~ "Phonological vs Control",
    "z_trial_n" ~ "z Trial number",
    "z_log_prime_duration" ~ "z Log prime duration",
    "z_log_target_duration" ~ "z Log target duration",
    "z_log_prime_rt" ~ "z Log prime RT",
    "z_isi" ~ "z ISI",
    .default = term
  )
}

target_fixed_effects <- brms::fixef(target_rt_brm) %>%
  as.data.frame() %>%
  rownames_to_column("term") %>%
  as_tibble() %>%
  filter(term != "Intercept") %>%
  mutate(
    draw_col = paste0("b_", term),
    parameter = map_chr(term, pretty_target_term),
    prob_gt_zero = map_dbl(draw_col, \(col) mean(target_draws[[col]] > 0)),
    `β` = sprintf("%.3f", Estimate),
    `SE` = sprintf("%.3f", Est.Error),
    CI = sprintf("[%.3f, %.3f]", Q2.5, Q97.5),
    `P(β > 0)` = case_when(
      prob_gt_zero > 0.999 ~ "> 0.999",
      prob_gt_zero < 0.001 ~ "< 0.001",
      TRUE ~ sprintf("%.3f", prob_gt_zero)
    )
  ) %>%
  select(parameter, `β`, `SE`, CI, `P(β > 0)`)

posterior_plot_draws <- target_draws %>%
  select(starts_with("b_")) %>%
  pivot_longer(everything(), names_to = "draw_col", values_to = "value") %>%
  mutate(term = str_remove(draw_col, "^b_")) %>%
  filter(term != "Intercept") %>%
  mutate(
    parameter = factor(
      map_chr(term, pretty_target_term),
      levels = rev(target_fixed_effects$parameter)
    )
  )

condition_hypothesis_strings <- c(
  "Morpho-Semantic vs. Morpho-Opaque" =
    "ptype_simpleMorphoSemantic - ptype_simpleMorphoOpaque > 0",
  "Morpho-Semantic vs. Pseudo-Morphological" =
    "ptype_simpleMorphoSemantic - ptype_simplePseudoMorphological > 0",
  "Morpho-Semantic vs. Phonological" =
    "ptype_simpleMorphoSemantic - ptype_simplePhonological > 0",
  "Morpho-Opaque vs. Pseudo-Morphological" =
    "ptype_simpleMorphoOpaque - ptype_simplePseudoMorphological > 0",
  "Morpho-Opaque vs. Phonological" =
    "ptype_simpleMorphoOpaque - ptype_simplePhonological > 0",
  "Pseudo-Morphological vs. Phonological" =
    "ptype_simplePseudoMorphological - ptype_simplePhonological > 0"
)

condition_hypotheses <- hypothesis(target_rt_brm, condition_hypothesis_strings)

contrast_draws <- tibble(
  contrast = names(condition_hypothesis_strings),
  draw_values = list(
    target_draws$b_ptype_simpleMorphoSemantic - target_draws$b_ptype_simpleMorphoOpaque,
    target_draws$b_ptype_simpleMorphoSemantic - target_draws$b_ptype_simplePseudoMorphological,
    target_draws$b_ptype_simpleMorphoSemantic - target_draws$b_ptype_simplePhonological,
    target_draws$b_ptype_simpleMorphoOpaque - target_draws$b_ptype_simplePseudoMorphological,
    target_draws$b_ptype_simpleMorphoOpaque - target_draws$b_ptype_simplePhonological,
    target_draws$b_ptype_simplePseudoMorphological - target_draws$b_ptype_simplePhonological
  )
) %>%
  unnest_longer(draw_values, values_to = "value")

condition_hypothesis_table <- condition_hypotheses$hypothesis %>%
  as.data.frame() %>%
  as_tibble() %>%
  mutate(contrast = names(condition_hypothesis_strings), .before = 1) %>%
  transmute(
    contrast,
    `β diff` = sprintf("%.3f", Estimate),
    `SE` = sprintf("%.3f", Est.Error),
    CI = sprintf("[%.3f, %.3f]", CI.Lower, CI.Upper),
    direction_prob = case_when(
      Post.Prob >= 0.5 & Post.Prob > 0.999 ~ "P(β > 0) > 0.999",
      Post.Prob >= 0.5 ~ sprintf("P(β > 0) = %.3f", Post.Prob),
      (1 - Post.Prob) > 0.999 ~ "P(β < 0) > 0.999",
      TRUE ~ sprintf("P(β < 0) = %.3f", 1 - Post.Prob)
    )
  )

contrast_plot_labels <- condition_hypotheses$hypothesis %>%
  as.data.frame() %>%
  as_tibble() %>%
  mutate(
    contrast = names(condition_hypothesis_strings),
    prob_label = case_when(
      Post.Prob >= 0.5 & Post.Prob > 0.999 ~ "P(β > 0) > 0.999",
      Post.Prob >= 0.5 ~ sprintf("P(β > 0) = %.2f", Post.Prob),
      (1 - Post.Prob) > 0.999 ~ "P(β < 0) > 0.999",
      TRUE ~ sprintf("P(β < 0) = %.2f", 1 - Post.Prob)
    )
  ) %>%
  transmute(
    contrast,
    prob_label,
    y = pmax(abs(CI.Lower), abs(CI.Upper)) * 1.08
  )
```

```{r target-rt-table}
kable(
  target_fixed_effects,
  align = c("l", "r", "r", "r", "r"),
  caption = "Posterior summary for the maximal Bayesian model of log target RTs. The intercept is omitted."
)
```

The coefficient table summarizes how each predictor shifts log target RT while holding the others constant. For the condition terms, the reference level is Control, so negative coefficients under $\beta$ column correspond to faster target responses relative to Control and positive coefficients correspond to slower target responses.

```{r target-rt-results-text, results='asis'}
strong_effects <- target_fixed_effects %>%
  mutate(prob_num = as.double(`P(β > 0)`)) %>%
  mutate(strong = pmax(prob_num, 1 - prob_num)) %>%
  arrange(desc(strong))

cat("The fixed-effect results can be read in terms of posterior direction probabilities under the P(β > 0) column. ")
cat("Values near 0.50 indicate substantial posterior uncertainty, whereas values near 1.00 or 0.00 indicate that the posterior mass is concentrated on one side of zero.\n\n")

for (i in seq_len(min(3, nrow(strong_effects)))) {
  direction_label <- if (strong_effects$prob_num[i] >= 0.5) {
    if (strong_effects$prob_num[i] > 0.999) {
      "P(β > 0) > 0.999"
    } else {
      paste0("P(β > 0) = ", sprintf("%.3f", strong_effects$prob_num[i]))
    }
  } else {
    if ((1 - strong_effects$prob_num[i]) > 0.999) {
      "P(β < 0) > 0.999"
    } else {
      paste0("P(β < 0) = ", sprintf("%.3f", 1 - strong_effects$prob_num[i]))
    }
  }
  cat(
    "- ",
    strong_effects$parameter[i],
    ": β = ",
    strong_effects$`β`[i],
    ", ",
    direction_label,
    ", CI ",
    strong_effects$CI[i],
    ".\n",
    sep = ""
  )
}
```

```{r target-rt-posterior-plot, fig.width=10, fig.height=4.2, fig.cap="Posterior distributions for fixed effects in the maximal log-target-RT model. Green slabs indicate facilitation (faster than Control); red slabs indicate inhibition (slower). The right column shows P(β < 0) in monospace, bolded when the posterior is strongly directional (> 0.89 or < 0.11). The intercept is omitted."}
# Compute per-parameter P(β < 0) and median direction
posterior_summary <- posterior_plot_draws %>%
  group_by(parameter) %>%
  summarise(
    prob_neg = mean(value < 0),
    median_val = median(value),
    .groups = "drop"
  ) %>%
  mutate(
    direction = ifelse(median_val < 0, "negative", "positive"),
    is_strong = prob_neg > 0.89 | prob_neg < 0.11,
    prob_label = case_when(
      prob_neg > 0.999 ~ "> 0.999",
      prob_neg < 0.001 ~ "< 0.001",
      TRUE ~ sprintf("%.3f", prob_neg)
    ),
    # Bold if strong
    prob_display = ifelse(is_strong,
      paste0("bold(\"", prob_label, "\")"),
      paste0("\"", prob_label, "\"")
    )
  )

# Join direction back to draws
plot_data <- posterior_plot_draws %>%
  left_join(posterior_summary %>% select(parameter, direction), by = "parameter")

# x position for the annotation column
x_annot <- max(plot_data$value, na.rm = TRUE) * 1.35

ggplot(plot_data, aes(x = parameter, y = value)) +
  geom_hline(yintercept = 0, linewidth = 0.4, linetype = "dashed", color = "grey55") +
  stat_slabinterval(
    aes(fill = direction, color = direction),
    side = "top",
    point_interval = median_qi,
    .width = c(0.66, 0.95),
    alpha = 0.7,
    point_color = "grey30",
    interval_color = "grey30",
    show.legend = FALSE
  ) +
  # Column header
  annotate(
    "text", 
    x = 9.5, 
    y = x_annot, 
    label = "underline('P(β < 0)')", # Note the quotes inside underline
    parse = TRUE,                   # This is critical!
    family = "mono", 
    size = 3.5, 
    hjust = 0.5
  ) +
  # P(β < 0) labels per parameter
  geom_text(
    data = posterior_summary,
    aes(x = parameter, y = x_annot,
        label = prob_label,
        fontface = ifelse(is_strong, "bold", "plain")),
    family = "mono", size = 3.3, hjust = 0.5,
    inherit.aes = FALSE
  ) +
  scale_fill_manual(values = c("negative" = "#66BB6A", "positive" = "#EF5350")) +
  scale_color_manual(values = c("negative" = "#4CAF50", "positive" = "#E53935")) +
  coord_flip(clip = "off") +
  labs(x = NULL, y = expression(beta)) +
  theme_classic(base_size = 12) +
  theme(plot.margin = margin(5.5, 60, 5.5, 5.5))
```

```{r target-rt-condition-contrasts-table}
kable(
  condition_hypothesis_table,
  align = c("l", "r", "r", "r", "r"),
  caption = "Posterior pairwise comparisons among the non-control prime conditions from `hypothesis()`."
)
```

The contrast table compares the non-control conditions directly. Each row tests whether the condition named on the left has a larger coefficient than the condition named on the right, and the direction-probability column states whether the posterior favors a positive or negative difference.

```{r target-rt-contrast-results-text, results='asis'}
contrast_results <- condition_hypothesis_table %>%
  mutate(prob_num = as.double(str_extract(direction_prob, "[0-9.]+$"))) %>%
  mutate(direction = if_else(str_detect(direction_prob, "P\\(β > 0\\)"), "positive", "negative"))

cat("These pairwise contrasts show whether the non-control prime conditions differ from each other, rather than only from Control.\n\n")

for (i in seq_len(min(3, nrow(contrast_results)))) {
  cat(
    "- ",
    contrast_results$contrast[i],
    ": estimated difference = ",
    contrast_results$`β diff`[i],
    ", ",
    contrast_results$direction_prob[i],
    ", CI ",
    contrast_results$CI[i],
    ".\n",
    sep = ""
  )
}
```

```{r target-rt-condition-contrasts-plot, fig.width=10, fig.height=3.5, fig.cap="Posterior distributions for pairwise differences among non-control prime conditions. Green slabs indicate a negative difference (left condition faster); red slabs indicate a positive difference (left condition slower). The right column shows P(Δβ < 0), bolded when strongly directional."}
# Prepare contrast draws with factor levels
contrast_draws_plot <- contrast_draws %>%
  mutate(
    contrast = factor(contrast, levels = rev(names(condition_hypothesis_strings)))
  )

# Compute per-contrast P(Δβ < 0) and direction
contrast_summary <- contrast_draws_plot %>%
  group_by(contrast) %>%
  summarise(
    prob_neg = mean(value < 0),
    median_val = median(value),
    .groups = "drop"
  ) %>%
  mutate(
    direction = ifelse(median_val < 0, "negative", "positive"),
    is_strong = prob_neg > 0.89 | prob_neg < 0.11,
    prob_label = case_when(
      prob_neg > 0.999 ~ "> 0.999",
      prob_neg < 0.001 ~ "< 0.001",
      TRUE ~ sprintf("%.3f", prob_neg)
    )
  )

# Join direction back
contrast_draws_plot <- contrast_draws_plot %>%
  left_join(contrast_summary %>% select(contrast, direction), by = "contrast")

# Annotation x position
x_annot_c <- max(contrast_draws_plot$value, na.rm = TRUE) * 1.35

ggplot(contrast_draws_plot, aes(x = contrast, y = value)) +
  geom_hline(yintercept = 0, linewidth = 0.4, linetype = "dashed", color = "grey55") +
  stat_slabinterval(
    aes(fill = direction, color = direction),
    side = "top",
    point_interval = median_qi,
    .width = c(0.66, 0.95),
    alpha = 0.7,
    point_color = "grey30",
    interval_color = "grey30",
    show.legend = FALSE
  ) +
  # Column header
  annotate(
    "text", 
    x = 6.5, 
    y = x_annot_c, 
    label = "underline(P(Delta*beta < 0))", # plotmath syntax
    parse = TRUE,                         # interprets the string as a formula
    family = "mono", 
    size = 3.5, 
    hjust = 0.5
  ) +
  # P(Δβ < 0) labels per contrast
  geom_text(
    data = contrast_summary,
    aes(x = contrast, y = x_annot_c,
        label = prob_label,
        fontface = ifelse(is_strong, "bold", "plain")),
    family = "mono", size = 3.3, hjust = 0.5,
    inherit.aes = FALSE
  ) +
  scale_fill_manual(values = c("negative" = "#66BB6A", "positive" = "#EF5350")) +
  scale_color_manual(values = c("negative" = "#4CAF50", "positive" = "#E53935")) +
  coord_flip(clip = "off") +
  labs(x = NULL, y = expression(Delta * beta)) +
  theme_classic(base_size = 12) +
  theme(plot.margin = margin(5.5, 60, 5.5, 5.5))
```
